\textbf{How open are hybrid journals included in transformative agreements?}

### 2024-02-19 analysis/manuscript.tex
starting discussions

For over two decades, hybrid open access journal publishing, which makes
some articles openly available while others remain behind a paywall, has
been discussed as a means for transitioning the subscription system to
full open access (Prosser, 2003). The idea was that when journals
increasingly publish open access articles, they could reduce revenues
from subscriptions, while libraries and funders could change their
funding models and shift expenditures from subscription to open access.
However, initial approaches, mainly based on publication fees, also
called article processing charges (APCs), did not contribute
substantially to a large open access uptake. In 2009, Springer reported
open access to 1\% of articles in hybrid journals (Dallmeier-Tiessen et
al., 2010). Other publishers also recorded a low uptake. In 2011, only
1-2\% of articles were open access (Björk, 2012), growing to around 4\%
between 2011 and 2013 (Laakso \& Björk, 2016).

With the introduction of central funding mechanisms for hybrid open
access publication fees in some European countries since 2012, an
increase in open access could be observed (Björk, 2017; Huang et al.,
2020; Jubb et al., 2017; Piwowar et al., 2018). For example, studying
university output, Robinson-Garcia et al. (2020) estimated a median
uptake of 7.1\% in the period 2014-2017. In particular, British (17\%),
Austrian (15\%) and Dutch (13\%) universities contributed to this trend.
However, this shift in funding policy towards hybrid open access also
added to the overall cost of publishing, which includes subscription
spending and the administrative efforts required to handle payments
(Pinfield et al., 2016). Moreover, established large commercial
publishers, which already dominated the publishing market (Larivière et
al., 2015), disproportionately benefited from hybrid open access funding
in comparison to full open access publishers (Butler et al., 2023; Jahn
\& Tullney, 2016; Shu \& Larivière, 2023).

As a consequence, libraries and their consortia began to develop
licensing strategies aimed at avoiding such `double dipping' scenarios,
in which well-established commercial publishers gain twice from reading
and publication fees, as well as to increase publisher-provided
immediate open access (Björk \& Solomon, 2014; Schimmer et al., 2015).
These considerations resulted into transformative agreements, which
cover a broad range of successfully negotiated contracts between library
consortia and publishers from mid-2010s onwards where institutional
spending for subscriptions are re-purposed for open access publishing
(Borrego et al., 2020; Hinchliffe, 2019). Transformative agreements seek
to control costs while allowing a transitional phase for publishing more
open access articles. Although transformative agreements mainly bundle
hybrid and subscription-only journals from commercial publishers, they
aim at a higher degree of transparency than previous big deals, where
contracts including payments were confidential (Bergstrom et al., 2014).

The introduction of transformative agreements aligns with funding policy
changes, such as the decision made by the cOAlition S, a consortium of
national funders including the European Commission, to no longer provide
financial support for individual publication fees when publishing in
hybrid journals. According ot its Plan S launched in 2018, the cOAlition
S funders only accept hybrid open access through transformative
agreements ``during a transition period that should be as short as
possible'' (Schiltz, 2018). Specifically, they agreed to support hybrid
open access solely through transformative agreements from 1 January
2021, until the end of 2024. Notably, the German Research Foundation
(DFG), despite not being part of cOAlition S, has also extended its
financial support for hybrid open access through transformative
agreements (Mittermaier, 2021). Previously, the DFG only provided
funding for full open access journals (Jahn \& Tullney, 2016).

By the end of 2023, many transformative agreements were implemented, but
outcomes were mixed. The ESAC Transformative Agreement
Registry\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}},
the largest source of disclosure, recorded more than 800 transformative
agreements, resulting in up to 900.000 open access articles published in
both full open access and hybrid journals according to the accompanying
ESAC Market Watch\footnote{\url{https://esac-initiative.org/market-watch/}}.
Library consortia reported increased open access volume, streamlined
payment and monitoring procedures, as well as extensive utilization of
open access options by the researchers they serve (Marques \& Stone,
2020; Parmhed \& Säll, 2023; Pinhasi et al., 2020). The ongoing
standardisation of transformative agreements contributed to improved
transparency in terms of contracts and publisher-provided article
metadata (Marques et al., 2019). However, with the growing trend toward
transformative agreements, continued reliance on big deals is perceived
as problematic, which might perpetuate market concentration (Butler et
al., 2023; Shu \& Larivière, 2023). Whether transformative agreements
lead to reduced pricing remains uncertain (Borrego, 2023) and a
substantial transition of hybrid journals towards full open access could
not be observed (Matthias et al., 2019; Momeni et al., 2021).
Furthermore, the focus on large commercial publishers might increase
inequality, because transformative agreements continue to target
institutions from high-income countries (Klebel \& Ross-Hellauer, 2023).
Besides, an editorial-board resignation raised concerns that publishers'
desire to maximize journal publication volume ``without regard to
quality'' is a consequence of transformative agreements (Rasmussen,
2023).

The controversies surrounding hybrid open access and transformative
agreements have led to varying policy conclusions. For instance, the
Association of Swedish Higher Education Institutions (Sveriges
universitets- och högskoleförbund, SUHF) recommended to only support
agreements for publishing in full open access journals \footnote{\url{https://www.su.se/english/news/open-access-need-to-move-away-from-transformative-agreements-1.683787}}.
Likewise, most cOAlition S funders will discontinue financial support
for transformative agreements from 2024 onwards (Liverpool, 2023). The
consortium also removed most hybrid journals from its Transformative
Journal program in 2023 due to their failure to meet self-defined open
access growth targets (Brainard, 2023). In contrast, the German DEAL
consortium signed a five-year transformative agreement with Elsevier
starting in 2024, while also renewing its contracts with Springer Nature
and Wiley until the end of 2028.

Despite these controversies around transformative agreements as a means
of transitioning journal publishing to full open access, there is
limited evidence available on the adoption of open access in hybrid
journals, and the extent to which this uptake can be attributed to
transformative agreements. Previous studies have focused on specific
countries (Haucap et al., 2021; Huang et al., 2020; Pölönen et al.,
2020; Taubert et al., 2023; Wenaas, 2022) or publisher portfolios
(Bakker et al., 2024; Fraser et al., 2023; Jahn et al., 2022; Momeni et
al., 2023; Pieper \& Broschinski, 2018), while related large-scale
studies relied on self-reported agreement data (Moskovkin et al., 2022),
or used APC pricing lists (Shu \& Larivière, 2023). Particularly, data
availability is a limiting factor when studying the impact of
transformative agreements (Bakker et al., 2024), because bibliometric
databases, even though many allow the retrieval of open access articles
in hybrid journals, do not directly attribute them to specific
transformative agreements.

The present study aims to address these limitations by combining
multiple openly available data sources to determine the open access
uptake in hybrid journals, while distinguishing between open access
through transformative agreements and other means. With this novel and
open approach, this study aims to answer the following questions:

\begin{itemize}
\tightlist
\item
  What was the number and proportion of open access articles in hybrid
  journals included in transformative agreements between 2018 and 2022?
\item
  How did transformative agreements impact open access uptake in hybrid
  journals?
\end{itemize}

@@

\hypertarget{methods-and-data}{%
\section{Methods and data}\label{methods-and-data}}

This study combines data from multiple publicly available data sources
as diagrammatically shown in Figure \ref{fig:data_workflow}. Initially,
transformative agreement data retrieved from the cOAlition S Journal
Checker Tool provided information about journal portfolios and
participating institutions. After identification of hybrid journals by
excluding full open access journals, Crossref served as the primary data
source for article-level metadata including Creative Commons (CC)
license information to indicate open access availability on publisher
websites. To determine open access articles published through
transformative agreements, first author affiliations from OpenAlex
(Priem et al., 2022) were subsequently linked to eligible institutions
according to the transformative agreement data.

@@

Because transformative agreements can include both full open access and
hybrid journals, the data were complemented with information about a
journal's open access status using multiple sources: the Directory of
Open Access Journals (DOAJ) downloaded on 12 December 2023\footnote{\url{https://doaj.org/csv}},
OpenAlex (November 2023) and the the Bielefeld list of GOLD OA journals
(Bruns et al., 2022). As shown in Figure \ref{fig:method_fig}A,
combining different data sources considerably extended the journal
matching to exclude full open access journals from transformative
agreements. In total, 3,439 full open access journals were excluded
based on ISSN matching. The overlap between the three data sources was
72\%. The Gold OA journals dataset alone added 176 journals, while the
DOAJ comprised 10 full open access journals not listed in either of the
other two sources. These full open access journals were mostly launched
in 2022.

@@

Open access articles in hybrid journals were identified through Creative
Commons (CC) license information in Crossref metadata. License
information relative to the ``accepted manuscript (AM)'' version were
not considered. Crossref was used for open access identification because
transformative agreements workflows generally require publishers to
deliver CC license information to this DOI registration agency (Geschuhn
\& Stone, 2017). Comparing Crossref license coverage with OpenAlex,
which re-uses open access evidence from Unpaywall, a widely used open
access discovery service that also parses journal websites for open
content licenses (Piwowar et al., 2018), highlighted ongoing challenges
to identify hybrid open access (Butler et al., 2023; Jahn et al., 2022;
Martín-Martín et al., 2018; Zhang et al., 2022). Here, 742,369 articles
with CC license were retrieved using Crossref, while 950,260 articles
were tagged as ``hybrid'' according to the OpenAlex November 2023
release, which was used throughout this study. The biggest differences
concerned articles published between 2018 and 2020. In 2022, however,
Crossref and OpenAlex open access numbers only differ slightly (249,511
records using Crossref vs.~255,344 in OpenAlex). Notable difference
could be furthermore observed among some publishers that presumably did
not provide CC license information to Crossref including AIP Publishing,
American Physiological Society and Emerald. Crossref license metadata
was more complete with regard to articles from the publisher Wiley and
American Chemical Society. Finally, inconsistent open access status
information in previous OpenAlex versions was observed (Jahn et al.,
2023). According to the OpenAlex release notes, fixing this issue was
still ongoing, which might also explain this discrepancy.

@@

Affiliation metadata about corresponding authors are crucial for the
planning and evaluation of transformative agreements, because they are
considered to be responsible to arrange open access publication (Borrego
et al., 2020; Geschuhn \& Stone, 2017; Schimmer et al., 2015). Here,
country and institutional affiliations were retrieved from OpenAlex.
However, because of low coverage in OpenAlex, this study focused on
first authors and their affiliations instead. First authors typically
contribute most to a paper and are often considered lead author research
papers (Larivière et al., 2016), and can be therefore assumed as a proxy
to measure to open access payments (Shu \& Larivière, 2023; Zhang et
al., 2022). Overall, around 90\% of studied articles had first author
affiliation metadata in OpenAlex, whereas the coverage of articles with
corresponding author information was around 54\%.

To assess the impact of transformative agreements to hybrid open access,
participating institutions from the Transformative Agreement Data dump,
which were crowd-sourced from the agreements and consortia that
successfully negotiated an agreement, were matched with first author
affiliations recorded by OpenAlex using the ROR-ID. The matching also
took into account the duration of an agreement according to the ESAC
registry. Upon inspection, Transformative Agreement Data did not cover
associated institutions comprehensively like university hospitals or
institutes of large research organisations like the Max Planck Society.
To improve the matching, Transformative Agreement Data was automatically
enriched with associated organisations using OpenAlex's institution
entity.

In total, the compiled data set consists of 8,922,146 articles published
in 12,857 hybrid journals between 2018 and 2022 (see Figure
\ref{fig:method_fig}B). Hybrid journals included in transformative
agreements represented 40\% of total global output over the same time
period according to Crossref, while full open access journals accounted
for 35\% of total article volume.

@@

Throughout this mostly automated data gathering and analysis process,
tools from the Tidyverse (Wickham et al., 2019) for the R programming
language (R Core Team, 2020) were used. The resulting data is openly
available through an R package, hoaddata, version 0.2.91 Following
Marwick et al. (2018), hoaddata contains not only the datasets used in
the data analysis. It also includes code used to compile the data by
connecting it the cloud-based Google Big Query data warehouse, where the
big scholarly data from Crossref, OpenAlex and Unpaywall were imported.
To increase the computational reproducibility, the R package was
automatically built using GitHub Actions, a continuous integration
service.

@@

Figure \ref{fig:results_overview}B highlights that the majority of open
access articles in hybrid journals were made available through
transformative agreements in 2021 and 2022, contributing 58\% of the
total open access article volume in 2022. However, there was also a
notable growth in open access provision through individual publication
fees, which increased from 4.1\% (n = 62,625) in 2018 to 6.3\% (n =
105,896). This suggests that publishers were able to gain equally from
individual and institutional open access publishing options.

Figure \ref{fig:results_overview}C depicts the substantial variations
among the hybrid journals included in transformative agreements in terms
of open access uptake. Although the median generally follows the trend
shown in Figure \ref{fig:results_overview}A, the farther stretch of
upper quartiles and whiskers over the years illustrates that an
increasing number of journals published an above-average proportion of
open access articles. In 2022, 25\% of hybrid journals (n = 2,576) had
an open access uptake of 29\%, and 6.6\% of journals (n = 744) provided
the majority of their articles under a Creative Commons license in the
same year. These journals were, on average, smaller (M = 75, SD = 186)
than those with an open access share below 50\% (M = 164, SD = 347).
Notable exception of large journals with an above-average open access
proportion were Physical Review D, a high-energy physics journal covered
by the SCOAP3 consortium that provided open access to 2,341 out of 4,074
articles in 2022, Astronomy and Physics (1,396 out of 2,230 articles in
2022 were open access), which shifted to a subscribe to open business
model for all accepted articles as of April 2022, and the Journal of
Fluid Mechanics (577 of 1,077 articles in 2022 were open access).

When comparing the impact of open access trough transformative
agreements across journals, it shows that for many journals these
agreements substantially contributed to the growth of open access over
the years (Figure \ref{fig:results_overview}D). Examples of such
journals include those with a scope on specific countries or regions,
where also transformative agreements were implemented. For instance, in
2022, the Germany-based journals Zeitschrift für Erziehungswissenschaft
and Zeitschrift für Politikwissenschaft, as well as the Scandinavian
Political Studies adressing the Nordic countries, achieved an overall
open access uptake of more than 90\% just through transformative
agreements. Despite the rise in transformative agreements, it is worth
noting that other means of publishing open access in hybrid journals
remained common. In total, 9,153 journals published open access articles
from authors affiliated with institutions without transformative
agreements in place, while 8,780 journals published at least one open
access article through a transformative agreement in the same year.

@@

Analysing hybrid open access across publishers between 2018 and 2022
reveals a large market concentration. Although 48 publishers offered
transformative agreements, the big three commercial publishers Elsevier,
Springer Nature, and Wiley, accounted for 49\% of total article volume
published (see Table \ref{tab:publisher_league_table}). Together, they
published 500,878 open access articles, representing 66\% of the open
access articles in hybrid journals. Elsevier, Springer Nature, and Wiley
made 243,891 articles open access in hybrid journals through
transformative agreements, resulting in an even larger market share of
74\%.

@@

However, there are notable differences among the three large publishers.
Although Elsevier published the largest volume of articles (n =
2,770,826, 34\%), it published a relatively low number of open access
articles, including those that can be associated with transformative
agreements. In contrast, Springer Nature and Wiley provided open access
to a larger proportion of their articles (13\% of Springer Nature
articles and 15\% of Wiley articles were open acccess), leading to
higher open access market shares (23\% Springer Nature resp. 23\%
Wiley). This difference between Elsevier on the one hand and Springer
Nature and Wiley on the other can be attributed to transformative
agreements, as the latter made the majority of their open access
articles available through such deals (Springer Nature 57\% resp. Wiley
55\%).

Figure \ref{fig:publisher_figure} takes a closer look into the growth of
hybrid open access across publishers by year with a focus on open
articles enabled by transformative agreements. Although all publishers
show a general long-term trend towards transformative agreements, Figure
\ref{fig:publisher_figure}A and B indicate that, in particular, Wiley's
has experienced a substantial increase in its open access share from
5.9\% (n = 11,628) in 2018 to 26\% (n = 53,503) in 2022, representing an
4.5-fold increase. In contrast, Elsevier's hybrid journals demonstrated
a more modest increase, from 3.3\% (n = 16,872) in 2018 to 10\% (n =
60,821) in 2022, which is a relatively low open access share compared to
the general trend. In 2018, Springer Nature had the largest open access
proportion among the three publishers of 8.4\% (n = 19,701), but
experienced a relatively slower growth, resulting in 18\% (n = 52,616)
of articles being open access in Springer Nature hybrid journals in
2022.

@@

The varying degrees of uptake of open access across the three major
publishers can be attributed to distinct approaches to transformative
agreements. Springer Nature, for example, began in 2015 offering
selected consortia, such as the Max Planck Society, the Swedish Bibsam
consortium, and the Finnish FinELib consortium, open access agreements
for its hybrid journal portfolio under the name Springer
Compact\footnote{\url{https://web.archive.org/web/20180414062853id_/http://www.liber2015.org.uk/wp-content/uploads/2015/03/Springer-Compact.pdf}}.
However, these agreements were not included in the data as they
concluded prior to the start of the transformative agreement data
collection in June 2021. Nonetheless, the results suggest the importance
of central agreements for Springer Nature's hybrid open access business
over the past five years (Figure 2B). In 2022, 66\% (n = 34,725) of open
access articles in Springer Nature hybrid journals were enabled through
transformative agreements. In the same year, 70\% (n = 37,316) of
Wiley's open access articles could be linked to transformative
agreements in 2022. In contrast, Elsevier published fewer than half of
its open access articles through transformative agreements (n = 32,627;
54\%).

The increasing trend towards transformative agreements can be also
observed at the journal-level (Figure \ref{fig:publisher_figure}C).
While no substantial differences between open access enabled through
transformative agreements and other revenue sources could observed
across Elsevier journals, the distribution of open access across
Springer Nature and Wiley hybrid journals indicates that the growth is
not limited to a few journals, but extends across the portfolio. In
particular, Wiley's upper quantile, which represents the top 25\% of
journals in terms of the proportion of open access articles from
transformative agreements, increased markedly from 13\% in 2020 to 31\%
in 2022. At the same time, the median proportion grew from 7.5\% to
19\%. It is interesting to note that a small but increasing number of
journals from these two publishers are providing open access to the
majority of articles through transformative agreements. Wiley recorded
68 and Springer Nature 102 hybrid journals with an open access share
above 50\% that could be solely attributed to transformative agreements.
Upon inspection, these journals were mainly society or local language
journals with a small yearly article volume.

>>>>

For over two decades, hybrid open access journal publishing, which makes
some articles openly available while others remain behind a paywall, has
been discussed as a means for transitioning the subscription system to
full open access (Prosser, 2003). The idea was that when journals
increasingly publish open access articles, they could reduce revenues
from subscriptions, while libraries and funders could change their
funding models and shift expenditures from subscription to open access.
However, initial approaches, mainly based on publication fees, also
called article processing charges (APCs), did not contribute
substantially to a large open access uptake. In 2009, Springer reported
open access to 1\% of articles in hybrid journals (Dallmeier-Tiessen et
al., 2010). Other studies also recorded a low uptake. In 2011, only
1-2\% of articles were open access (Björk, 2012), growing to around 4\%
between 2011 and 2013 (Laakso \& Björk, 2016).

With the introduction of central funding mechanisms for publication fees
in some European countries since 2012, an increase in hybrid open access
could be observed (Björk, 2017; Huang et al., 2020; Jubb et al., 2017;
Piwowar et al., 2018). For example, studying university output,
Robinson-Garcia et al. (2020) estimated a median uptake of 7.1\% in the
period 2014-2017. In particular, British (17\%), Austrian (15\%) and
Dutch (13\%) universities contributed to this trend. However, this shift
in funding policy towards hybrid open access also added to the overall
cost of publishing, which includes subscription spending and the
administrative efforts required to handle payments (Pinfield et al.,
2016). Moreover, established large commercial publishers, which already
dominated the publishing market (Larivière et al., 2015),
disproportionately benefited from hybrid open access funding in
comparison to full open access publishers (Butler et al., 2023; Jahn \&
Tullney, 2016; Shu \& Larivière, 2023).

As a consequence, libraries and their consortia began to develop
licensing strategies aimed at avoiding such `double dipping' scenarios,
in which well-established commercial publishers gain twice from reading
and open access publishing, as well as to increase publisher-provided
immediate open access (Björk \& Solomon, 2014; Schimmer et al., 2015).
These considerations resulted into transformative agreements, which
cover a broad range of contracts between library consortia and
publishers from mid-2010s onwards where institutional spending for
subscriptions and open access publishing are considered together
(Borrego et al., 2020; Hinchliffe, 2019). Transformative agreements seek
to control costs while allowing a transitional phase for publishing more
open access articles. Similar to big deals, transformative agreements
mainly bundle hybrid and subscription-only journals from commercial
publishers, but aim at a higher degree of transparency than previous big
deals, where contracts including payments were confidential (Bergstrom
et al., 2014).

The introduction of transformative agreements aligns with funding policy
changes, such as the decision made by the cOAlition S, a consortium of
national funders including the European Commission, to no longer provide
financial support for individual publication fees when publishing in
hybrid journals. According ot its Plan S launched in 2018, the cOAlition
S funders only accept hybrid open access through transformative
agreements ``during a transition period that should be as short as
possible'' (Schiltz, 2018). Specifically, they agreed to support hybrid
open access only through transformative agreements from 2021, until the
end of 2024. Notably, the German Research Foundation (DFG), despite not
being part of cOAlition S, has also extended its financial support for
hybrid open access through transformative agreements (Mittermaier,
2021). Previously, the DFG only provided funding for full open access
journals (Jahn \& Tullney, 2016).

By the end of 2023, many transformative agreements were implemented, but
interim outcomes are mixed. The ESAC Transformative Agreement
Registry\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}},
the largest source of disclosure, recorded more than 800 transformative
agreements, resulting in up to 900.000 open access articles published in
both full open access and hybrid journals according to the accompanying
ESAC Market Watch\footnote{\url{https://esac-initiative.org/market-watch/}}.
Library consortia reported increased open access volume, streamlined
payment and monitoring procedures, as well as extensive utilization of
open access options by the researchers they serve (Marques \& Stone,
2020; Parmhed \& Säll, 2023; Pinhasi et al., 2020). The ongoing
standardisation of transformative agreements contributed to improved
transparency in terms of contracts and publisher-provided article
metadata (Marques et al., 2019; Pinhasi et al., 2021). However, with the
growing trend toward transformative agreements, continued reliance on
big deals is perceived as problematic, because it perpetuate market
concentration (Butler et al., 2023; Shu \& Larivière, 2023). Whether
transformative agreements lead to reduced pricing remains uncertain
(Borrego, 2023) and a substantial transition of hybrid journals towards
full open access could not be observed (Matthias et al., 2019; Momeni et
al., 2021). The focus on large commercial publishers might also increase
inequality, because transformative agreements focus on pay to publish
open access mainly targets institutions from high-income countries,
furthering a questionable journal prestige culture (\textbf{budapest?}).
Besides, an editorial-board resignation raised concerns that publishers'
desire to maximize journal publication volume ``without regard to
quality'' is a consequence of transformative agreements (Rasmussen,
2023).

The controversies surrounding hybrid open access and transformative
agreements have led to varying policy conclusions. For instance, the
Association of Swedish Higher Education Institutions (Sveriges
universitets- och högskoleförbund, SUHF) recommended to only support
agreements for publishing in full open access journals \footnote{\url{https://www.su.se/english/news/open-access-need-to-move-away-from-transformative-agreements-1.683787}}.
Likewise, most cOAlition S funders will discontinue financial support
for transformative agreements by the end of 2024 (Liverpool, 2023). The
consortium also removed most hybrid journals from its Transformative
Journal program in 2023 due to publishers' failure to meet self-defined
open access growth targets (Brainard, 2023). In contrast, the German
DEAL consortium signed a five-year transformative agreement with
Elsevier starting in 2024, while also renewing its contracts with
Springer Nature and Wiley until the end of 2028.

Despite these controversies around transformative agreements as a means
of transitioning journal publishing to full open access, there is
limited evidence available on the adoption of open access in hybrid
journals, and the extent to which this uptake can be attributed to
transformative agreements. Previous studies have focused on specific
countries (Haucap et al., 2021; Huang et al., 2020; Pölönen et al.,
2020; Taubert et al., 2023; Wenaas, 2022) or publisher portfolios
(Bakker et al., 2024; Fraser et al., 2023; Jahn et al., 2022; Momeni et
al., 2023; Pieper \& Broschinski, 2018), while large-scale studies
relied on self-reported agreement data (Moskovkin et al., 2022), or used
APC pricing lists (Shu \& Larivière, 2023). Particularly, data
availability is a limiting factor when studying the impact of
transformative agreements (Bakker et al., 2024), because bibliometric
databases, even though many allow the retrieval of open access articles
in hybrid journals, do not directly attribute them to specific
transformative agreements.

The present study aims to address these limitations by combining
multiple openly available data sources to determine the open access
uptake in hybrid journals, while distinguishing between open access
through transformative agreements and other means. With this novel and
open approach, this the first large-scale analysis aims to answer the
following questions:

\begin{itemize}
\tightlist
\item
  What was the number and proportion of open access articles in hybrid
  journals in transformative agreements between 2018 and 2022?
\item
  To what extent did institutions with a transformation agreement
  contribute to the adoption of open access in hybrid journals?
\end{itemize}

@@

\hypertarget{methods}{%
\section{Methods}\label{methods}}

This study combines data from multiple publicly available data sources
as diagrammatically shown in Figure \ref{fig:data_workflow}. Initially,
transformative agreement data retrieved from the cOAlition S Journal
Checker Tool\footnote{\url{https://www.coalition-s.org/blog/enabling-accurate-results-within-the-journal-checker-tool/}}
provided information about journal portfolios and participating
institutions. After identification of hybrid journals by excluding full
open access journals, Crossref served as the primary data source for
article-level metadata including Creative Commons (CC) license
information to indicate open access availability on publisher websites.
To determine open access articles published through transformative
agreements, first author affiliations from OpenAlex (Priem et al., 2022)
were subsequently linked to eligible institutions according to the
transformative agreement data. In the following, these steps are
described in more detail.

@@

Because transformative agreements can include both full open access and
hybrid journals, the data were complemented with information about a
journal's open access status using multiple sources: the Directory of
Open Access Journals (DOAJ) downloaded on 12 December 2023\footnote{\url{https://doaj.org/csv}},
OpenAlex (November 2023) and the the Bielefeld list of GOLD OA journals
(Bruns et al., 2022). As shown in Figure \ref{fig:method_fig}A,
combining different data sources considerably extended the journal
matching. In total, 3,439 full open access journals were excluded based
on ISSN matching. The overlap between the three data sources was 72\%.
The Gold OA journals dataset alone added 176 journals, while the DOAJ
comprised 10 full open access journals not listed in either of the other
two sources. These full open access journals were mostly launched in
2022.

@@

Open access articles in hybrid journals were identified through Creative
Commons (CC) license information in Crossref metadata. License
information relative to the ``accepted manuscript (AM)'' version were
not considered. Crossref was used for open access identification because
transformative agreements workflows generally require publishers to
deliver CC license information to this DOI registration agency (Geschuhn
\& Stone, 2017). Comparing Crossref license coverage with OpenAlex,
which re-uses open access evidence from Unpaywall, a widely used open
access discovery service that also parses journal websites for open
content licenses (Piwowar et al., 2018), highlighted ongoing challenges
to identify hybrid open access (Butler et al., 2023; Jahn et al., 2022;
Martín-Martín et al., 2018; Zhang et al., 2022). Here, 742,369 articles
with CC license were retrieved using Crossref, while 950,260 articles
were tagged as ``hybrid'' according to the OpenAlex November 2023
release, which was used throughout this study. The biggest differences
concerned articles published between 2018 and 2020. In 2022, however,
Crossref and OpenAlex open access numbers only differ slightly (249,511
records using Crossref vs.~255,344 in OpenAlex). Notable difference
could be furthermore observed among some publishers that presumably did
not provide CC license information to Crossref including AIP Publishing,
American Physiological Society, Emerald and the Royal Society. Crossref
license metadata was more complete with regard to articles from the
publisher Wiley and American Chemical Society. Finally, inconsistent
open access status information in previous OpenAlex versions was
observed (Jahn et al., 2023). After reporting it to OpenAlex, according
to the release notes, fixing this issue was still ongoing, which might
also explain part of the discrepancy.

@@

Affiliation metadata about corresponding authors are crucial for the
planning and evaluation of transformative agreements, because they are
considered to be responsible to arrange open access publication (Borrego
et al., 2020; Geschuhn \& Stone, 2017; Schimmer et al., 2015). Here,
country and institutional affiliations were retrieved from OpenAlex.
However, because of low coverage in OpenAlex, this study focused on
first authors and their affiliations instead. First authors typically
contribute most to a paper and are often considered lead author research
papers (Larivière et al., 2016). Related studies therefore assumed first
authors as a proxy to measure to open access payments and the impact of
transformative agreement (Haucap et al., 2021; Shu \& Larivière, 2023;
Zhang et al., 2022). Overall, around 90\% of studied articles had first
author affiliation metadata in OpenAlex, whereas the coverage of
articles with corresponding author information was around 54\%.

To assess the impact of transformative agreements to hybrid open access,
participating institutions from the Transformative Agreement Data dump,
which were crowd-sourced from the agreements and consortia that
successfully negotiated an agreement, were matched with first author
affiliations recorded by OpenAlex using the ROR-ID. The matching also
took into account the duration of an agreement according to the ESAC
registry. Upon inspection, Transformative Agreement Data did not cover
associated institutions comprehensively like university hospitals or
institutes of large research organisations like the Max Planck Society.
To improve the matching, Transformative Agreement Data was automatically
enriched with associated organisations using OpenAlex's institution
entity data.

In total, the compiled data set consists of 8,922,146 articles published
in 12,857 hybrid journals between 2018 and 2022 (see Figure
\ref{fig:method_fig}B). Hybrid journals in transformative agreements
represented 40\% of total global output over the same time period
according to Crossref, while full open access journals recorded 35\%.

@@

Throughout this mostly automated data gathering and analysis process,
tools from the Tidyverse (Wickham et al., 2019) for the R programming
language (R Core Team, 2020) were used. The resulting data is openly
available through an R data package, hoaddata, version 0.2.91 Following
Marwick et al. (2018), hoaddata contains not only the datasets used in
the data analysis. It also includes code used to compile the data by
connecting it the cloud-based Google Big Query data warehouse, where the
big scholarly data from Crossref, OpenAlex and Unpaywall were imported.
To increase the computational reproducibility, the data aggregation
through hoaddata was automatically carried out using GitHub Actions, a
continuous integration service.

@@

Figure \ref{fig:results_overview}B highlights that the majority of
hybrid open access was made available through transformative agreements
in 2021 and 2022, contributing 58\% of the total open access article
volume in 2022. However, there was also a notable growth in open access
provision through other means, most likely publication fees being not
invoiced through transformative agreements, which increased from 4.1\%
(n = 62,625) in 2018 to 6.3\% (n = 105,896).

Figure \ref{fig:results_overview}C depicts the substantial variations
among the hybrid journals included in transformative agreements in terms
of open access uptake. Although the median generally follows the trend
shown in Figure \ref{fig:results_overview}A, the farther stretch of
upper quartiles and whiskers over the years illustrates that an
increasing number of journals published an above-average proportion of
open access articles. In 2022, 25\% of hybrid journals (n = 2,576) had
an open access uptake of 29\%, and 6.6\% of journals (n = 744) provided
the majority of their articles under a Creative Commons license in the
same year. These journals were, on average, smaller (M = 75, SD = 186)
than those with an open access share below 50\% (M = 164, SD = 347).

When comparing the impact of open access trough transformative
agreements across journals, it shows that for many journals these
agreements substantially contributed to the growth of open access over
the years (Figure \ref{fig:results_overview}D). Despite the rise in
transformative agreements, it is worth noting that other means of
publishing open access remained common across the investiagted hybrid
journals. In total, 9,153 journals published open access articles from
authors affiliated with institutions without transformative agreements
in place, while 8,780 journals published at least one open access
article through a transformative agreement in the same year.

@@

Analysing hybrid open access across publishers between 2018 and 2022
reveals a large market concentration. Although 48 publishers offered
transformative agreements, the big three commercial publishers Elsevier,
Springer Nature, and Wiley accounted for 49\% of total article volume
published (see Table \ref{tab:publisher_league_table}). Together, they
published 500,878 or 66\% of open access articles in hybrid journals.
Elsevier, Springer Nature, and Wiley made 243,891 articles open access
in hybrid journals through transformative agreements, resulting in an
even larger market share of 74\%.

@@

However, there are notable differences among the three big publishers.
Although Elsevier published the largest volume of articles (n =
2,770,826, 34\%), it recorded a comparable low number of open access
articles, including those that can be associated with transformative
agreements. In contrast, Springer Nature and Wiley provided open access
to a larger proportion of their articles (13\% of Springer Nature
articles and 15\% of Wiley articles were open acccess), leading to
higher open access market shares (23\% Springer Nature resp. 23\%
Wiley). This difference between Elsevier on the one hand and Springer
Nature and Wiley on the other can be attributed to transformative
agreements, as the latter made the majority of their open access
articles available through such deals (Springer Nature 57\% resp. Wiley
55\%).

Figure \ref{fig:publisher_figure} takes a closer look into the growth of
hybrid open access across publishers by year with a focus on open
articles enabled by transformative agreements. Although all publishers
show a general long-term trend towards transformative agreements, Figure
\ref{fig:publisher_figure}A and B indicate that, in particular, Wiley
experienced a substantial increase in its open access share from 5.9\%
(n = 11,628) in 2018 to 26\% (n = 53,503) in 2022, representing an
4.5-fold increase. In contrast, Elsevier's hybrid journals demonstrated
a more modest increase, from 3.3\% (n = 16,872) in 2018 to 10\% (n =
60,821) in 2022, which is a relatively low open access share compared to
the general trend. In 2018, Springer Nature had the largest open access
proportion among the three publishers of 8.4\% (n = 19,701), but
experienced a relatively slower growth, resulting in 18\% (n = 52,616)
of articles being open access in Springer Nature hybrid journals in
2022.

@@

The varying degrees of adoption of open access across the three major
publishers can be attributed to distinct approaches to transformative
agreements. Springer Nature, for example, began in 2015 offering
selected consortia, such as the Max Planck Society, the Swedish Bibsam
consortium, and the Finnish FinELib consortium, open access agreements
for its hybrid journal portfolio under the name Springer
Compact\footnote{\url{https://web.archive.org/web/20180414062853id_/http://www.liber2015.org.uk/wp-content/uploads/2015/03/Springer-Compact.pdf}}.
However, these agreements were not included in the data as they
concluded prior to the start of the transformative agreement data
collection in June 2021. Nonetheless, the results demonstrate the
importance of agreements for Springer Nature's hybrid open access
business over the past five years (Figure 2B). In 2022, 66\% (n =
34,725) of open access in Springer Nature hybrid journals were enabled
through transformative agreements. In the same year, 70\% (n = 37,316)
of Wiley's open access articles could be linked to transformative
agreements in 2022. In contrast, Elsevier published fewer than half of
its open access articles through transformative agreements (n = 32,627;
54\%).

The increasing trend towards transformative agreements can be also
observed at the journal-level (Figure \ref{fig:publisher_figure}C).
While no substantial differences between open access enabled through
transformative agreements and other revenue sources could observed
across Elsevier's portfolio, the distribution of open access across
Springer Nature and Wiley hybrid journals indicates that the growth is
not limited to a few journals, but extends across the portfolio. In
particular, Wiley's upper quantile, which represents the top 25\% of
journals in terms of the proportion of open access articles from
transformative agreements, increased markedly from 13\% in 2020 to 31\%
in 2022. At the same time, the median proportion grew from 7.5\% to
19\%. It is interesting to note that a small but increasing number of
journals from these two publishers are providing open access to the
majority of articles through transformative agreements. Wiley recorded
68 and Springer Nature 102 hybrid journals with an open access share
above 50\% that could be solely attributed to transformative agreements.
Upon inspection, these journals were mainly society or local language
journals with a small yearly article volume.
